Machine Learning Models for the Segmentation of Clinical Exchange

### 2018-03-08 paper-amia-2018/manuscript.tex
update paper; working on LSTM model

\title{Machine Learning Models for the Segmentation of E-coaching Text}

@@

\section*{Methods}
\subsection*{\textit{Data collection}}
The experimental dataset for this work was constructed from the 49 e-coaching sessions, which include a total of 3,138 segmented and annotated MI behaviors. Each session contains an MI intervention involving patient-provider communications in email. To filter out noise from the dataset, non-ascii characters are removed and then applied stemming to obtain a general form of word from different word representations, such as ``eating'', ``eats'', and ``eat''. We formulate the text segmentation task into a binary classification, as shown in Figure~\ref{fig:classifier}. An intervention session with email exchange is given as the input, it is partitioned into adjacent word pairs by sliding them. Each pair of them classified into either of the two categories: ``boundary'' and ``not boundary''. The text is segmented at the position, where an adjacent word pair classified into ``boundary'' class. If all pairs of word classified into ``not boundary'', the text is treated as one about a single MI behavior. Totaly, we obtained 95,421 word pairs, which include 3,138 ``boundary'' and 92,283 ``not boundary'' instances.

@@

For the experiment, we utilized three type of features including word (textual feature), topic, and punctuation mark. Each word represented in a binary format, where 1 indicates the appearance of the word and 0 for absence. Topics are considered as features since topic models are very effective\cite{kotov2015interpretable,hashimoto2016topic,lu2016modeling} to represent text documents. In this paper, we exploit the Labeled LDA model\cite{kotov2015interpretable} and represent each word in a vector of 2 topics, where the number of topics is experimentally determined by the model performance. Punctuation mark containing one of the symbols \{`.', `,', `!', `?', `:', `;', `-'\} is also employed as feature. This is one of the most important features as they indicate the boundary of a sentence, clause, and phrase.

@@

Table~\ref{tab:result_not_boundary} summarizes the performance of NB, SVM, KNN, and RNN models for detecting ``not boundary'' class in e-coaching text. We observed that performance is remarkably high in ``not boundary'' class compared to boundary detection. Similar to boundary detection, KNN consistently outperforms over all other methods but obtains 22.40\%, 50.07\%, and 36.26\% higher precision, recall, and F1-measure for contextual features; and 20.85\%, 33.96\%, and 27.47\% higher precision, recall, and F1-measure for combined features compared to boundary class. However, RNN demonstrates the lowest performance and exhibits 0.981 and 0.983 precision with 0.986 and 0.987 F1-measure for LSTM and GRU, respectively. Impact of additional features is also consistent with ``not boundary'' classification. Results show that F1-measure increases by 0\%, and 0.1\% for SVM and KNN models although decreases by 0.1\% for NB. \\

@@

\section*{Discussion}
This study is the first large-scale efforts to evaluate the segmentation of e-coaching text. Experimental results indicate that KNN is the best model among all machine learning methods considered for this study. KNN achieved 0.986 F1-measure in overall, 0.779 and 0.993 F1-measures for detecting ``boundary'' and ``not boundary'', respectively. The robust performance of KNN provides the evidence that machine learning models are capable to learn information from the email exchange. Although the domain of this study was intentionally quite small, we believe that our study is not limited to the e-coaching domain, and it can be successfully applied to other domain as well.

@@

In this paper, results are reported by each class to avoid confusion about the overall model performance. In addition, standard metrics: precision, recall, and F1-measure were used to eliminate doubt about the model performance because accuracy is misleading for imbalance dataset. AUC values are also outlined due to its effectiveness in measuring the quality of binary classifiers for imbalanced datasets\cite{hu2015kernelized}, which are demonstrated by the ROC curves in Figure~\ref{fig:roc-curves}. NB shows the highest AUC values, achieved 0.978 for both cases while provides lowest classification results except RNN. On the other hand, KNN and SVM exhibit 0.972 and 0.835 AUCs when only textual features are used; and 0.959 and 0.844 AUCs when a combination of textual, topic and punctuation mark features are used. Finally, RNN demonstrates lowest AUC values among all machine learning models, achieved AUC values 0.704 and 0.740 for LSTM and GRU, respectively.

We observed worst results of RNN, in particular, LSTM and GRU for the text segmentation. We believe that RNN performed poorly because it has a large set of weights which required longer sequence with a large set of data for both classes. In this study, maximum length of the input sequence was 3 and we utilized 3,138 examples of boundary case which failed to achieve good results. While the performance of RNN is poor, GRU performs better than LSTM which was observed in other previous study\cite{chung2014empirical}.

Punctuation mark plays an important role in segmentation boundary detection, and large numbers of errors were encountered by the false positive of boundary identification. For example, ``xxxxx xxxxxx'' was segmented after the second sentence where a punctuation mark was encountered. Similarly, additional information is the common reason for the classified original segment into multiple segments. For instance, ``xxxx xxxx xxxx'' was segmented after second sentence where third sentence supports the MI code confirmed by first two sentences.

Our proposed approach is novel for the segmentation of e-coaching text because previous studies mainly focus on the segmentation of text into sections, headers, and sentences in other medical domain. However, this study segmented email exchange into groups of MI behaviors. This work will significantly reduce the amount of resource and time required to segment email text manually. Furthermore, this paper can help to annotate each segment automatically by building a new classifier, which will accelerate the pace of finding best communication strategies to develop an effective MI intervention for healthy eating.

As our future work, we plan to evaluate our approach on other datasets involves in discourse analysis. We also plan to use a combination of machine learning and natural language approaches to improve model performance. For example, part-of-speech tagging and distance from the boundary of the sentence might significantly enhance the classification performance. The limitation of this study is that e-coaching text is collected from a single medical institute; formatting, style, and email segment can be different in other settings.

\section*{Conclusion}
Segmentation of e-coaching text is an integral part of developing an automated e-coaching intervention. Although several studies have done for the segmentation of clinical text into sections and sentences, none of them are used for the segmentation of text into groups of MI behaviors in the setting of discourse analysis with email under the principle of motivational interviews. In this paper, we compared the performance of machine learning models for the task of segmentation of e-coaching text. We found out that k-nearest neighbor provides the best performance for the segmentation of text in terms of all performance metrics. Manual segmentation of e-coaching data is very resource-intensive and time-consuming task, which can significantly decrease the time and effort required to develop an effective behavioral intervention. Our proposed methods can help to identify individual text segments, which can be annotated directly with a classification model. This approach will also help for developing fully automated e-coaching and accelerate the pace of identifying effective communication strategies linked to healthy eating.

\section*{Acknowledgments}
This study was supported by a grant from the National Institutes of Health, NIDDK R21DK108071, Carcone and Kotov, MPIs. We would like to thank the student assistants in the Department of Family Medicine and Public Health Sciences at Wayne State University School of Medicine for their help in developing the training dataset by manually segmented the e-coaching text.

>>>>

\title{Machine Learning Models for the Segmentation of Clinical Exchange}

@@

\section*{Methods}
\subsection*{\textit{Data collection}}
The experimental dataset for this work was constructed from 49 e-coaching sessions, which include a total of 3,138 segmented and annotated MI behaviors. Each session represents an MI intervention delivered via email. To filter out noise from the dataset, non-ascii characters are removed and then applied stemming to obtain a general form of word from different word representations, such as ``eating'', ``eats'', and ``eat''. We formulate the text segmentation task into a binary classification, as shown in Figure~\ref{fig:classifier}. Clinical exchange is given as the input, it is partitioned as adjacent word pairs by sliding them. Each pair is classified into either ``boundary'' or ``not boundary'' class. The original text is segmented at the position, where an adjacent word pair classified into ``boundary'' class. If all adjacent pairs of a block of text classified into ``not boundary'', the whole block of text is then treated as one segment about a single MI behavior. Totaly, we obtained 95,421 word pairs, which include 3,138 ``boundary'' and 92,283 ``not boundary'' instances.

@@

For the experiment, we utilized three type of features including word (textual feature), topic, and punctuation mark. Each word represented in a binary format, where 1 indicates the appearance of the word and 0 for absence. Topics are considered as features since topic models are very effective\cite{kotov2015interpretable,hashimoto2016topic,lu2016modeling} to represent text documents. In this paper, we exploit a topic model named Labeled LDA\cite{ramage2009labeled} model, where topics are corresponds to labels. The generative process of Labeled LDA drawing the multinomial topic distributions over vocabulary for each topic. In our experiment, we represent each word in a vector of 2 topics, where the number of topics is experimentally determined by the model performance\cite{kotov2015interpretable}. Punctuation mark containing one of the symbols \{`.', `,', `!', `?', `:', `;', `-'\} is also employed as feature. This is one of the most important features as they indicate the boundary of a sentence, clause, and phrase.

@@

Table~\ref{tab:result_not_boundary} summarizes the performance of NB, SVM, KNN, and RNN models for detecting ``not boundary'' class in e-coaching text. We observed that performance is remarkably high in ``not boundary'' class compared to boundary detection which is expected because 96.71\% instances are from ``not boundary'' class. Similar to boundary detection, KNN consistently outperforms over all other methods but obtains 22.40\%, 50.07\%, and 36.26\% higher precision, recall, and F1-measure for contextual features; and 20.85\%, 33.96\%, and 27.47\% higher precision, recall, and F1-measure for combined features compared to boundary class. However, RNN demonstrates the lowest performance and exhibits 0.981 and 0.983 precision with 0.986 and 0.987 F1-measure for LSTM and GRU, respectively. Impact of additional features is also consistent with ``not boundary'' classification. Results show that F1-measure increases by 0\%, and 0.1\% for SVM and KNN models although decreases by 0.1\% for NB. \\

@@

\section*{Discussion}
This study is the first efforts to evaluate the automatic segmentation of e-coaching text. Experimental results indicate that KNN is the best model among all machine learning methods considered for this study. KNN achieved 0.986 F1-measure in overall, 0.779 and 0.993 F1-measures for detecting ``boundary'' and ``not boundary'', respectively. The robust performance of KNN provides the evidence that machine learning models are capable to learn information from clinical exchange. Although the domain of this study was intentionally quite small, we believe that our study is not limited to the e-coaching domain, and it can be successfully applied to other domain as well.

@@

In this paper, results are reported by each class to avoid confusion about the overall model performance. In addition, standard metrics: precision, recall, and F1-measure were used to eliminate doubt about the model performance because accuracy is misleading for imbalance dataset. AUC values are also outlined due to its effectiveness in measuring the quality of binary classifiers for imbalanced datasets\cite{hu2015kernelized}, which are demonstrated by the ROC curves in Figure~\ref{fig:roc-curves}. NB shows the highest AUC values, achieved 0.978 for both cases while provides lowest classification results except RNN. On the other hand, KNN and SVM exhibit 0.972 and 0.835 AUCs when only textual features are used; and 0.959 and 0.844 AUCs when a combination of textual, topic and punctuation mark features are used. Finally, RNN demonstrates lowest AUC values among all machine learning models, achieved AUC values 0.704 and 0.740 for LSTM and GRU, respectively. The conclusion drawn from the ROC curves, also confirmed the robustness and superiority of KNN model for the segmentation of clinical exchange.

We observed worst results of RNN, in particular, LSTM and GRU for the text segmentation. We believe that RNN performed poorly because it has a large set of weights which required longer sequence with a large set of data for all labels. In this study, maximum length of the input sequence was 3 and we utilized 3,138 examples of boundary case which failed to achieve good results. While the performance of RNN is poor, GRU performs better than LSTM which was observed in other previous study\cite{chung2014empirical}.

Punctuation mark plays an important role in segmentation boundary detection, and large numbers of errors were encountered by the false positive of boundary identification. For example, a text block ``A1 A2 A3. B1 B2 B3. C1 C2 C3 C4.'' can be segmented at position A3, B3, and C4 where a punctuation mark was encountered. Similarly, additional information is the common reason for the classified original segment into multiple segments. For instance, the above text block can be segmented at position B3 and C4 because third sentence (C) only supports the MI code confirmed by first two sentences (A and B).

Our proposed approach is novel for the segmentation of e-coaching text because previous studies mainly focus on the segmentation of text into sections, headers, and sentences in other medical domain. However, this study segmented clinical exchange into groups of MI behaviors which will significantly reduce the amount of resource and time required to segment clinical exchange manually. Furthermore, these segmentation models could be integrated with auto coding classifiers to build a software package of automatic coding procedure of clinical exchange.

The limitation of this study is that e-coaching text is collected from a single medical institute; formatting, style, and email segment can be different in other settings. Therefore, there is a need to replicate the experiments with different data sets. As our future work, we plan to evaluate our approach on other datasets involves in discourse analysis. We also plan to use more relevant features to improve model performance. For example, part-of-speech tagging\cite{hasan2016feedback} and distance from the begining and end of the current sentence, might significantly enhance the classification performance.

\section*{Conclusion}
Segmentation of e-coaching text is an integral part of developing an automated e-coaching intervention. Although several studies have done in clinical interventions, they are limited by the qualitative coding of clinical interactions. In addition, previous studies in medical domain, mainly segmented clinical text into sections and sentences, none of them are used for the segmentation of text into groups of MI behaviors in the setting of discourse analysis with email under the principle of motivational interviews. In this paper, we compared the performance of machine learning models for the task of segmentation of e-coaching text. We found out that k-nearest neighbor provides the best performance for the segmentation of text in terms of all performance metrics. Manual segmentation of e-coaching data is very resource-intensive and time-consuming task, which can significantly decrease the time and effort required to develop an effective behavioral intervention. Our proposed methods can help to identify individual text segments, which can be annotated directly with a classification model. This approach will also help for developing fully automated e-coaching and accelerate the pace of identifying effective communication strategies.

\section*{Acknowledgments}
This study was supported by a grant from the National Institutes of Health, NIDDK R21DK108071, Carcone and Kotov, MPIs. We would like to thank research staff and student assistants in the Department of Family Medicine and Public Health Sciences at Wayne State University School of Medicine for their help in preparing the training dataset.
